The Rockerverse: Images, Packages, and Applications for Containerization with R

### 2020-01-23 manuscript.Rmd
shorten liftr section and move dockerfile stuff

## Capture and create environments [@nuest]
\label{envs}

Several second order R packages attempt to make the process of creating Docker images and using containers for specific tasks, such as running tests or rendering reproducible reports, easier.

**\pkg{dockerfiler}** ([https://github.com/ColinFay/dockerfiler/](https://github.com/ColinFay/dockerfiler/)) ...

@@

**\CRANpkg{liftr}** ([https://nanx.me/liftr/](https://nanx.me/liftr/)) \citep{liftr2019} aims to solve the problem of persistent reproducible reporting in statistical computing.
Currently, the R Markdown format and its backend compilation engine _knitr_ offer a _de facto_ standard for creating dynamic documents \citep{xie2018}.
However, the reproducibility of such content authoring environments is often limited to individual machines --- it is not easy to replicate the system environment (libraries, R versions, R packages) where the document was compiled.
This issue becomes even more serious when it comes to collaborative document authoring and creating large-scale document building services.
_liftr_ solves this reproducibility problem by bringing Docker to the game.
In essence, _liftr_ helps R Markdown users create and manage Docker containers for rendering the documents, thus make the computations utterly reproducible across machines and systems.

On implementation, with no side effects, _liftr_ extended and introduced new metadata fields to R Markdown, allowing users to declare the dependencies for rendering the document.
_liftr_ parses such fields and generates a `Dockerfile` for creating Docker containers.
_liftr_ then helps render the document inside the created Docker container.
This workflow is summarized in Figure \ref{figure:liftr}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{liftr-workflow}
  \caption{The liftr workflow for rendering containerized R Markdown documents.}
  \label{figure:liftr}
\end{figure}

_knitr_ and R Markdown are used as the template engine to generate the `Dockerfile`.
Features such as caching container layers for saving image build time, automatic housekeeping for fault-tolerant builds, and Docker status check are supported by _liftr_.
Four RStudio addins are also offered by _liftr_ to allow push-button compilation of documents and provide better IDE integrations.

Three basic principles are followed to design the _liftr_ package since its inception.

1. Continuous reproducibility.
Continuous integration, continuous delivery, and continuous deployment are well-accepted practices in software engineering.
Similarly, it is believed by the authors of _liftr_ that ensuring computational reproducibility means a continuous process instead of creating static data/code archives or a one-time deal.
Specifically, the software packages used in data analysis should be upgraded regularly in a manageable way.
Therefore, _liftr_ supports specifying particular versions of package dependencies, while users are encouraged to always use the latest version of packages (without a version number) by default.

2. Document first.
Many data analysis workflows could be wrapped as either R packages or dynamic documents.
In _liftr_, the endpoint of dynamic report creation is the focus of containerization, because this offers more possibilities for organizing both computations and documentation.
Users are encouraged to start thinking from the visible research output from the first day.

3. Minimal footprint.
R Markdown and Docker are already complex software systems.
Making them work together seamlessly can be complicated.
Therefore, API designs such as function arguments are simplified while being kept as expressive and flexible as possible.

In summary, _liftr_ tries to redefine the meaning of computational reproducibility by offering system-level reproducibility for data analysis.
It provided a practical way for achieving it --- a new perspective on how reproducible research could be done in reality.
Further, sharing system environments for data analysis also becomes extremely easy, since users only need to share the R Markdown document (with a few extra metadata fields), and compile them with _liftr_.
As an example, _liftr_ demonstrated its advantage for R Markdown-based computational workflow orchestration, by effortlessly containerizing 18 complex _Bioconductor_ workflows in the DockFlow project (https://dockflow.org) in 2017.

One of the key tool of Docker are `Dockerfile`s, which can be thought of as "recipes" used to build a specific image.
But building these files can feel like a workflow break as it demands to open a separate file, and / or to write a small wrapper around R's `write()` function to add elements to the file.
At the same time, doing it by hand prevents from using programming language for what they are good for: iteration, and automation.
Iteration and automation for `Dockerfile`s is the very reason behind [`dockerfiler`](https://github.com/ColinFay/dockerfiler), an R package designed for building `Dockerfile`s straight from R.

For example, the **[`golem`](https://github.com/ThinkR-open/golem)** package makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building production-grade Shiny applications and deploying them.
The reason behind this is that `dockerfiler`, on top of being scriptable from R, can leverage all the tools available in R to parse a `DESCRIPTION` file, to get system requirements, to list dependencies, versions, etc.

>>>>

## Capture and create environments
\label{envs}

Several second order R packages attempt to make the process of creating Docker images and using containers for specific tasks, such as running tests or rendering reproducible reports, easier.
While authoring and managing an environment with Docker by hand is possible and feasible for experts\footnote{See, e.g., this tutorial by RStudio on how to manage environments and package versions and to ensure deterministic image builds with Docker: \href{https://environments.rstudio.com/docker}{https://environments.rstudio.com/docker}.}, the following examples show the power of automation when environments become too cumbersome or acquiring the skills is not possible in due course.

**\pkg{dockerfiler}** ([https://github.com/ColinFay/dockerfiler/](https://github.com/ColinFay/dockerfiler/)) ...
At the same time, doing it by hand prevents from using programming language for what they are good for: iteration, and automation.
Iteration and automation for `Dockerfile`s is the very reason behind [`dockerfiler`](https://github.com/ColinFay/dockerfiler), an R package designed for building `Dockerfile`s straight from R.
For example, the **[`golem`](https://github.com/ThinkR-open/golem)** package makes an heavy use of `dockerfiler` when it comes to creating the `Dockerfile` for building production-grade Shiny applications and deploying them.
The reason behind this is that `dockerfiler`, on top of being scriptable from R, can leverage all the tools available in R to parse a `DESCRIPTION` file, to get system requirements, to list dependencies, versions, etc.

@@

**\CRANpkg{liftr}** aims to solve the problem of persistent reproducible reporting in statistical computing\citep[\href{https://nanx.me/liftr/}{https://nanx.me/liftr/}, ][]{liftr2019}.
Currently, the R Markdown format and its backend compilation engine _knitr_ offer a _de facto_ standard for creating dynamic documents \citep{xie2018}.
The reproducibility authoring environments is often limited to individual machines and becomes an issue for collaborative documents and large-scale platforms for processing documents.
\CRANpkg{liftr} helps R Markdown users to create and manage Docker containers for rendering the documents.
It introduced new fields to the R Markdown document header, allowing users to declare the dependencies, including versions, for rendering the document.
The package then generates a `Dockerfile` from this metadata and provides a utitility function to render the document inside a Docker container, e.g., `render_docker("foo.Rmd")`.
An RStudio addins even allows compilation of documents with a single push of a button.
\CRAKpkg{liftr} makes the document the main and sole workflow control file and the only file that needs to be shared between collaborators for consistent workflow environments, e.g. demonstrated in the DockFlow project (https://dockflow.org).
